Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-12 Paper/application.tex
Added missing references. Improved related work and conc. Now 12 pages

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that originally supports only basic string operations such as concatenation and regular constraints. CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.

By incorporating our SFT-based modeling of replacement, we develop CertiStrR—an extended string solver that supports two types of string replacement: (1) replacing the first occurrence of a substring matching a given pattern, and (2) replacing all occurrences of substrings matching a given pattern. The patterns can be either constant strings or regular expressions.

@@

\paragraph{Transducers vs. Replacement Operations.}  
Although both transducers and replacement operations can express string transformations for security-related verification, transducers are strictly more expressive because they can capture dependencies between the input pattern and the replacement string. In contrast, all replacement operations in SMT-LIB treat the pattern and the replacement string as independent. For example, in the SFT shown in Figure~\ref{fig:html-sanitizer-sft}, a transition may be defined over an interval such as $[\texttt{a-z}]$, and the output function computes its output directly based on the input character. Replacement operations, however, cannot express such dependencies.
Because the replacement string is not a function.

@@

This subsection discusses the development effort required for our certified SFT formalization. While SFTs constitute the central contribution of this paper, they are not sufficient on their own to support the applications we target. A more comprehensive symbolic framework for automata libraries is necessary. Beyond the existing automata library supported by CertiStr \cite{cpp/KanLRS22}, which already includes operations such as SFA product and SFA concatenation, we needed to develop additional \emph{certified} SFA operations from scratch. These include $\varepsilon$SFAs, translation from $\varepsilon$SFAs to SFAs, translation from SFAs to Symbolic Deterministic Finite Automata (SDFA), and negation of SFAs. Table~\ref{tab:abstract_impl} summarizes the certified development effort that was carried out specifically for this work.

>>>>

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that supports only basic string operations such as concatenation and regular constraints. CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.

By incorporating our SFT-based modeling of replacement, we develop CertiStrR—an
extended string solver that supports two types of string replacement: (1)
replacing the first occurrence of a substring matching a given pattern, and (2)
replacing all occurrences of substrings matching a given pattern, i.e., either 
a constant string or a regular expression.

@@

\paragraph{Transducers vs. Replacement Operations.}  
Although both transducers and replacement operations can express string
transformations for security-related verification, transducers are strictly more
expressive because they can capture dependencies between the input pattern and
the replacement string. In contrast, all replacement operations in SMT-LIB treat
the pattern and the replacement string as independent. For example, in the SFT
in Figure~\ref{fig:html-sanitizer-sft}, a transition may be defined over
an interval such as $[\texttt{a-z}]$, and the output function computes its
output directly based on the input character. Replacement operations, however,
cannot express such dependencies since the replacement string is not a function.

@@

%This subsection discusses the development effort required for our certified SFT formalization. 
While SFTs constitute the central contribution of this paper, they are not sufficient on their own to support the applications we target. A more comprehensive symbolic framework for automata libraries is necessary. Beyond the existing automata library supported by CertiStr \cite{cpp/KanLRS22}, which already includes operations such as SFA product and SFA concatenation, we needed to develop additional \emph{certified} SFA operations from scratch. These include $\varepsilon$SFAs, translation from $\varepsilon$SFAs to SFAs, translation from SFAs to Symbolic Deterministic Finite Automata (SDFA), and negation of SFAs. Table~\ref{tab:abstract_impl} summarizes the certified development effort that was carried out specifically for this work.

### 2025-09-12 Paper/conclusion.tex
Added missing references. Improved related work and conc. Now 12 pages

In this paper, we presented the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL. Our framework offers flexible interfaces that enable diverse applications, featuring support for $\varepsilon$-transitions on both inputs and outputs, as well as extensibility to various Boolean algebras through a refinement framework.

To demonstrate the practical utility of our formalization, we applied SFTs to model string sanitization functions commonly used in web applications and showed how to verify the correctness of these sanitizers. Furthermore, we developed CertiStrR, an extension of CertiStr \cite{cpp/KanLRS22}, which adds support for string replacement operations. We evaluated CertiStrR on benchmarks from SMT-LIB 2025 \cite{smtlib_benchmarks}, and the experimental results confirm both the efficiency and effectiveness of our approach.

>>>>

We have presented the first formalization of symbolic finite transducers (SFTs)
and their most crucial algorithms in Isabelle/HOL. Our framework offers 
flexible interfaces that enable diverse applications, featuring support for 
$\varepsilon$-transitions on both inputs and outputs, as well as extensibility to various Boolean algebras through a refinement framework.
To demonstrate the practical utility of our formalization, we applied formalized
SFTs for analyzing sanitization for web applications, as well as certified 
string solving for complex constraints involving equality, concatenation,
regular constraints, and replaceall, for which we confirm the efficiency
and effectiveness of our approach on SMT-LIB 2025 benchmarks
\cite{smtlib_benchmarks}.
%We envision that our work will benefit the string solving community by 
%providing a formal foundation for string solvers development.
%from SMT-LIB 2025 \cite{smtlib_benchmarks}, and the experimental results confirm both the efficiency and effectiveness of our approach.
%applied SFTs to model string sanitization functions commonly used in web applications and showed how to verify the correctness of these sanitizers. Furthermore, we developed CertiStrR, an extension of CertiStr \cite{cpp/KanLRS22}, which adds support for string replacement operations.

### 2025-09-12 Paper/relatedwork.tex
Added missing references. Improved related work and conc. Now 12 pages

\emph{String Solving.} A significant application of our certified transducer
framework is in string constraint solving, a field that has seen intensive
research development over the past decade. While our work provides formal
verification guarantees, there exists a rich ecosystem of non-certified string
solvers, which has in the last few years led to a standardized SMT-LIB 2.6/2.7 
unicode string theory and annual competition. SMT-LIB compatible string 
solvers include Z3 \cite{z3}, Z3-alpha \cite{z3alpha}, Z3-Noodler 
\cite{noodler,noodler-int,noodler-len, noodler-tool}, Z3str3RE \cite{z3str3re},
cvc5 \cite{cvc5}, Z3str4 \cite{z3str4}, Trau \cite{trau0,trau-tool,trau2}, and 
OSTRICH \cite{ostrich}. Recent SMT competition over string theory has revealed
superiority of automata-based methods (based on restrictions of symbolic
automata). For this reason,
%: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
%As more and more 
%Numerous bugs have been uncovered in existing string solvers \cite{DBLP:conf/cav/BlotskyMBZKG18} and SMT solvers \cite{Mansur20} 
our work will benefit the community by providing a formal 
foundation for string solvers development.

\emph{Certified SMT Solvers.} Beyond string theories, certification efforts in SMT solving have extended to other domains. For example, the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of interactive theorem proving in certified SMT solver development. The work by Mansur et al. \cite{verified-verifying} define a formal semantics for SMT-LIB string
to check compatibility between solvers and SMTLIB standard.

>>>>

\emph{String Solving.} A significant application of our certified transducer
framework is in string constraint solving, a field that has seen intensive
research development over the past decade. While our work provides formal
verification guarantees, there exists a rich ecosystem of non-certified string
solvers, which has in the last few years led to a standardized SMT-LIB 2.6/2.7 
unicode string theory and annual competition. SMT-LIB compatible string 
solvers include Z3 \cite{Z3}, Z3-alpha \cite{z3alpha}, Z3-Noodler 
\cite{noodler,noodler-int,noodler-len, noodler-tool}, Z3str3RE \cite{z3str3re},
cvc5 \cite{cvc5}, Z3str4 \cite{z3str4}, Trau \cite{Z3-trau,trau-tool}, and 
OSTRICH \cite{ostrich,CHL+19,pacmpl/ChenFHHHKLRW22}. 
Recent SMT competition over string theory has revealed
superiority of automata-based methods (based on restrictions of symbolic
automata). 
We envision that our work will benefit the string solving community by 
providing a formal foundation for string solvers development.
%: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
%As more and more 
%Numerous bugs have been uncovered in existing string solvers \cite{DBLP:conf/cav/BlotskyMBZKG18} and SMT solvers \cite{Mansur20}

\emph{Certified SMT Solvers.} Certification efforts 
for SMT (other than string theories) also exist. For example, the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of interactive theorem proving in certified SMT solver development. The work by Mansur et al. \cite{verified-verifying} define a formal semantics for SMT-LIB string
to check compatibility between solvers and SMTLIB standard.
